The interactions of rational, pragmatic agents\\ lead to efficient language structure and use

### 2019-01-31 paper/paper.tex
update objective function -- integrate comments from NDG about emphasis on novelty of objective and our framing.

\begin{abstract}
Languages display a diverse set of distributional regularities such as
the relation between a word's frequency and rank in a corpus, the
distribution of dependency lengths, or the presence of lexical
properties such as ambiguity. We discuss a framework for studying how
these properties emerge from in-the-moment interactions of rational,
pragmatic speakers and listeners. Our work takes notions of
lexicon-level efficiency as a starting point, connecting these ideas to
notions of conversational-level efficiency. To do so, we derive an
objective function for measuring the communicative efficiency of
linguistic systems and then examining the behavior of this objective in
a series of simulations focusing on the communicative function of
ambiguity in language. These simulations suggest that rational pragmatic
agents will produce communicatively efficient systems and that
interactions between such agents provide a framework for examining
efficient properties of language more broadly.

@@

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows -- how does having \textbf{property x} make using language
\(\ell\) more or less useful for communication? This reframing provides
an opportunity to study both language \emph{structure} (as Zipf
primarily did) or \emph{use} in terms of efficiency. For example, Regier
et al. (2015) showed that languages appear to organize semantic domains
(form-meaning mappings) in maximally efficient ways. Likewise,
Piantadosi et al. (2011) argued that lexical ambiguity is an efficient
property of any communication system when communication is
contextualized.\par

Zipf's original work and subsequent functionalist projects describe a
varied set of linguistic properties and human behaviors. Common to many
is a characterization of a fundamental effort-asymmetry underlying
everyday communication. Simply put, what is ``hard'' for a speaker is
likely different from what is ``hard'' for a listener. Zipf
characterized this as follows -- purely from the standpoint of speaker
effort, an optimal language \(\ell_{speaker}^*\) would tend toward a
vocabulary of just a single, low-cost word. Given such a language, the
full set of potential meanings would be conveyed using only that word,
i.e. \(\ell_{speaker}^*\) would be fully ambiguous and all possible
meanings would need to be disambiguated by a listener. From the
standpoint of listener effort, an optimal language \(\ell_{listener}^*\)
would map all possible meanings to distinct words, removing a listener's
need to disambiguate. In this example, speaker effort is related to
\emph{production cost} and listener effort to \emph{understanding or
disambiguation cost}. Clearly natural languages fall between the two
extremes of \(\ell_{speaker}^*\) and \(\ell_{listener}^*\). Zipf
proposed that the particular lexicon-level properties he observed were a
result of these competing forces -- the pressure to jointly minimize
speaker and listener effort.\par

But how does such optimization take place? The example given by Zipf
(1949), describes local, communicative interactions in terms of a
\textit{reference game}. Speakers intend to refer to some object in the
world \(m\). They choose some utterance \(u\) to transmit this intended
meaning, \(u \rightarrow m\). The listener attempts to reconstruct this
intended meaning given the transmitted utterance, \(m \rightarrow u\).
Other functionalist projects have assumed this basic reference game
setting (Piantadosi et al., 2011; Regier et al., 2015) and this
simplification of the communicative act has proven productive in
theoretical (Ferrer-i-Cancho, 2018), simulation-based (Kirby, Griffiths,
\& Smith, 2014) and empirical explorations (Hawkins, Franke, Smith, \&
Goodman, 2018) of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to
speakers' pragmatic reasoning in the moment. Under a Gricean treatment
of pragmatics, speakers and listeners are assumed to follow a set of
conversational maxims in which they cooperate to transfer information
(Grice, 1975). These maxims appear to emerge from efficiency concerns,
however (Horn, 1984). A primary target of the current work is to
formalize this connection, by providing a framework for deriving
system-level efficiencies from local, interaction behavior. Our claim is
that to understand an ``efficient'' property of a system we need to
consider \emph{how} that property is \emph{used} efficiently.
Additionally, any quantitative description of ``efficiency'' should be a
function of both the system and users.\par

In the current work we provide a case-study for the kind of framing
presented above. We choose a property of languages that could, in
principle, vary freely, but shows strong regularities across languages.
The explanadum is why this regularity holds. We examine lexical
ambiguity as our property, extending ideas by Piantadosi et al. (2011).
We define a novel measure of efficiency that depends on the
interactional behavior of interlocutors. For our model of speakers and
listeners we adopt the Rational Speech Act (RSA) framework, which is
nicely attested in experimental data on pragmatic interaction. With
these basic ingredients we have a measure of efficiency given language
and speaker-listener agents. Finally, we show that the property of
interest is prevalent in languages that optimize this measure of
efficiency and also how this property is used efficiently during local
interaction. The primary contribution of this work is to show how the
reference game frameowrk, in combination with formal models of
communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

Zipf's example of optimal speaker- and listener-languages took the form
of a reference game. We adopt that formulation here, assuming these
communication games as our basic unit of analysis. In this setting,
speakers and listeners are aware of a set of objects \(M\), which we
will refer to as \emph{meanings} and are knowledgeable about the set of
possible signals \(U\) (\emph{utterances}) that can be used to refer to
a given meaning (see Figure 1). Utterances may have different relative
costs, operationalized via a prior over utterances \(P(U)\). Similarly,
meanings differ in the relative degree to which they need to be talked
about, operationalized as a prior over meanings
\(P(M)\)\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}.
We consider a set of contexts \(C\) with an associated prior \(P(C)\).
Contexts encode different meaning priors, formalized as separate
conditional distributions over meanings e.g.
\(p(M|C=c_i) \neq p(M|C=c_j)\). Finally, we consider a set of
communicative events \(e \in E\) where \(<u, m> = e\) is a tuple of
utterance, meaning pairs. Note that in the case of single context
\(|C|=1\) we simply have a prior over meanings \(p(M)\), a
simplification we will use in introducing our objective function.\par

@@

A language \(\ell\) defines the set of semantic mappings between
utterance and meanings. For example, Figure 1 contains four utterances
\(U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}\).
The boolean matrix describes the literal semantics of the language
\(\ell\). Note that will consider language to ``contain ambiguous
items'' if there is at least one utterance \(u \in U\) which can apply
to multiple meanings (i.e.
\(|[[u_i]]| > 1\))\footnote{We use double brackets $[[\dots]]$ to represent denotation}.
For example, in Figure 1 both the words ``blue'' and ``green'' are
ambigous so we would say that \(\ell\) contains ambiguity.

@@

The Rational Speech-act framework (RSA) is computational-level theory of
pragmatic language use. RSA can be understood as a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context. For this reason, we adopt RSA as our
representational framework to operationalize Gricean (rational and
pragmatic) speaker-listeners (Grice, 1975). We refer the reader to
Section 1 of our supplementary materials for a more detailed description
of basic RSA speaker-listener definitions, and to Goodman \& Frank
(2016) for an overview of its appliations.\par

An RSA \emph{speaker agent} defines a conditional distribution over
utterances, mapping from intended meanings \(M\) to utterances \(U\)
using \(\ell\) in a given context \(c\). That is, a speaker defines
\(P_{speaker}(u|m; \ell)\). We will use \(S(u|m; \ell)\) as short-hand
throughout. A \emph{listener agent} defines a conditional distribution
over meanings, mapping from utterances \(U\) to meanings \(M\) using
\(\ell\) in a given context \(c\). We will use \(L(m|u; \ell)\) as
shorthand. Note that both speakers and listeners can induce joint
distributions over the set of all events \(E\), although, importantly,
these distributions may differ:
\[P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)\]
\[P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)\]

@@

Zipf (1949) proposed that the particular distributional properties found
in natural language emerge from competing speaker and listener
pressures. We operationalize this objective in equation (1) -- the
efficiency of a linguistic system \(\ell\) being used by speaker and
listener agents \(S\) and \(L\) is the sum of the expected speaker and
listener effort to communicate over all possible communicative events
\(E\). Note that to simplify the introduction of this objective we
assume there is just a single context (\(|C|=1\)). In this case
\(p(C=c)=1\) and there is only distribution over meanings
(\(p(m|c) = p(m)\)). We provide a more detailed description of this
derivation in the supplementary materials, both with and without context
\(C\).\par

@@

We assume that speaker effort is the negative log probability of an
utterance (surprisal) -- intuitively, the number of bits needed to
encode the utterance \(u\). This particular formalization of
speaker-cost is general enough to accommodate a range of instantiations
that might in theory be related to production difficult via articulation
effort, cognitive effort related to lexical access, etc (Bennett \&
Goodman, 2018).\par

\[\text{speaker effort} = -log_2(p(u))\]

We assume listener effort is the negative log probability of the
listener's conditional \(-log_2(L(m|u))\). This operationalization of
listener effort is intuitively related to existing work in sentence
processing in which word difficulty during comprehension is proportional
to surprisal (Hale, 2001; R. Levy, 2008).

\[\text{listener effort} = -log_2(L(m|u; \ell))\]

Importantly, we assume that events \(e = <u, m>\) are sampled according
the to following generative model -- some object occurs in the world
with probability \(p(M=m)\), the speaker attempts to refer to that
object by sampling from her conditional distribution \(S(u|m)\) (i.e.
\(e \sim p(m)S(u|m)\)). From the ingredients highlighted above, it is
possible to derive the following objective between the speaker and
listener joint distributions (see SI 2.2 for a complete derivation).

\begin{equation}
\begin{split}
  = -\mathbb{E}_{P_{speaker}}[log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}

From an information-theoretic perspective this objective is intuitive:
cross-entropy gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, this is the difference between the joint distribution assumed
by the speaker \(P_{speaker}\) and listener \(P_{listener}\). In other
words, an ``efficient'' language \(\ell\) minimizes the distance between
what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to
handle various forms of ambiguity; lexical, syntactic, among others
(Wasow, Perfors, \& Beaver, 2005). The ubiquity of this property,
however, has been argued to provide evidence that langauges are not been
optimized for
communication\footnote{Chomsky (2002) famously claimed "If you want to make sure that we never misunderstand one another, for that purpose language is not well designed, because you have such properties as ambiguity." Note that this analysis is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$ we described earlier and define in terms of equation 2 in SI 2.2 .}
(Chomsky, 2002).

@@

We provide a toy example of their argument here. Say we have two objects
in the world \(m_1\) and \(m_2\), and two languages \(\ell_1\) and
\(\ell_2\). In language \(\ell_1\), the low-cost utterance \(u_1\) can
be used to refer to both \(m_1\) and \(m_2\) (i.e.
\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\) and
\([[u_2]]_{\ell_1} = \emptyset\)). With language \(\ell_2\), however,
the low cost utterance \(u_1\) can only refer to \(m_1\) and a
higher-cost utterance \(u_2\) can refer to \(m_2\) (i.e.
\([[u_1]]_{\ell_2} = \{m_1\}\) and \([[u_2]]_{\ell_2} = \{m_2 \}\)).
While it is cheaper for a speaker to use \(\ell_1\) (because she can
always use a low-cost utterance), it is more difficult for a listener
(because \(u_1\) is ambiguous). If context isn't disambiguating then we
might prefer \(\ell_2\) to \(\ell_1\). But, if context is disambiguating
then the speaker can use \(u_1\) to refer to either \(m_1\) or \(m_2\)
and \(\ell_1\) should be preferred.\par

@@

\caption[Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases]{Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only objectives.}\label{fig:plot-optimal-langs}
\end{figure*}
\end{CodeChunk}

@@

We adopt the framework above, and (following Piantadosi et al. 2011),
add \textit{context} (\(C\)). The authors argue that mapping two
meanings \(m_1\) and \(m_2\) to a single utterance \(u_1\) is useful
when they can be disambiguated \emph{in context}. In our case we
consider a context \(c_i \in C\) to specify a permutation of the meaning
probabilities. That is, \(p(m|c_1) \neq p(m|c_2)\), when there are two
contexts \(|C| = \{c_1, c_2\}\).This in turn, leads to an update to our
linguistic efficiency objective as we now would like to consider the
average speaker-listener effort over all contextualized communicative
events (see section 3 SI).\par

We proceed by generating languages with different amounts of contextual
support (varying the size of \(|C|\)). We search the space of languages,
asessing if the ones which minimize our objective contain ambiguity. If
context leads to more efficient commnication, then optimal languages
should be more likley to be ambiguous as the amount of context
increases.\par

@@

We conduct \(N=2000\) simulations. For each simulation we enumerate the
set of \emph{valid} languages in which \(|U|=|M|=4\) (recall that \(U\)
is our set of utterances and \(M\) our set of meanings). Note that a
language \(\ell \in L\) is ``valid'' so long as each possible meaning in
\(m \in M\) can be referred to by at least one form \(u \in U\) (every
column of \(\ell\) has some non-zero assignment) and each form maps to
at least one meaning (every column has some non-zero
assignment)\footnote{For the current set of simulations we consider languages in which there are four or five true values -- in a "four" language ambiguity is not possible, but in a "five" language there are ambiguous mappings.}.
For a given simulation the goal is to find the language \(\ell^*\) which
minimizes our cross-entropy objective and then check to see if that
language contains ambiguity.\par

Recall that language efficiency is both a function of the particular
semantic mappings induced by that language, the speaker and listener
agents (\(S\) and \(L\)), as well as the utterance (\(P(U)\)), meaning
(\(P(M)\)), and context priors (\(P(C)\)). Rather than assume particular
structure for our utterance, meaning, and context prior distributions,
for each simulation we generate \(P(U) \sim \text{Dir}(1, |U|)\),
\(P(M|C=c) \sim \text{Dir}(1, |M|)\) (a separete conditional
distribution over meanings for each context \(c\)), and
\(P(C) \sim \text{Dir}(1, |C|)\) where \(\text{Dir}(1, k)\) specifies
the uniform Dirichlet distribution over a \(k\)-dimensional probability
vector.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on our objective. To do so we
consider four conditions with \(n=500\) simulations each (that is, 500
unique sets of \(\{P(U), P(M|C), P(C)\}\). Our first is a
\emph{single-context} condition (\(|C|=1\)) -- there is a only a single
context describing \(P(M)\). Our second condition contains two-contexts
(\(|C| = 2\)) -- we consider efficiency under both \(P(M|C=c_1)\) as
well as \(P(M|C=c_2)\). The third and fourth condition correspond
accordingly with \(|C|=3\) and \(|C| = 4\), respectively.\par

@@

For comparison, we also examine properties of optimal languages under
two additional objectives. Zipf (1949) proposed that the optimal speaker
language \(\ell_{speaker}^*\) would only optimize speaker effort and the
optimal listener language \(\ell_{listener}^*\) would only optimize
listener effort. We define these objectives using the first and second
half of equation 1 (see SI section 2.2. for more detail).\par

@@

Figure 1. plots the proportion of optimal languages under each
objective, as a function of context condition. The red line shows that
as the number of contexts increases, so does the probability that the
optimal language \(\ell^*_{cross}\) under our speaker-listener
cross-entropy objective contains ambiguity. We also plot the proportion
of optimal speaker-only \(\ell^*_{speaker}\) and listener-only
\(\ell^*_{listener}\) languages that contain ambiguity. In line with
Zipf's predictions, if languages are designed only to minimize speaker
effort then optimal languages will assign lower-cost forms multiple
meanings (all of the speaker-optimal languages contained ambiguity).
Likewise, if languages are designed only to minimize listener effort
then ambiguity should always be avoided (none of the listener-optimal
languages contained ambiguity).\par

We explored the degree to which ambiguity improved efficiency during
communication between rational, pragmatic agents. We showed that as
contextual information increased, the optimal languages under our
Cross-Entropy objective were more likely to have ambiguous lexical
items. Importantly, we did not see this effect under alternative
objectives which only emphasized speaker- or listener effort alone.\par

While our results indicate that ambiguity is an efficient property of
contextualized language these simulations assumed that speaker-listener
agents could always disambiguate items in context. That is, in our four
conditions both agents had perfect knowledge of the relevant conditional
distributions (\(P(M|C)\)). This assumption may be too strong for
describing much of day-to-day communication -- we seldom begin a
conversation with perfect knowledge of the current context of a
conversation. We explore a case in which the listener has imperfect
knowledge of context at the start of the conversation in Simulation
2.\par

@@

We shift time-scales in Simulation 2 -- instead of examining the
objective landscape of a space of languages, we now consider a single
language that contains ambiguity and examine how that property is used
over a discourse. If context is not informative with respect to meaning
early in a discourse (position \(i\)), but is at later positions at
position (position \(i+k\)), then the speaker should avoid using
ambigous material early \(i\), only using it \(k\) steps later in the
conversation. In this toy example we can consider ``context'' as
analogous to a ``topic'' of conversation. Intuitively, we can imagine a
scenario in which a reader is beginning a newspaper article. While they
may have some knowledge about the article's topic (perhaps from the
title), they may not have complete knowledge of its contents, including
the persons or events involved. In this setting, using an ambiguous
pronoun early (position \(i\)) in the article can lead to
misunderstanding without sufficient context to support disambiguation.
But if by position \(i+k\) there is sufficient contextual information it
may be efficient for the writer to use ambiguous material.\par

@@

>>>>

\begin{abstract}
Languages display a diverse set of distributional regularities such as
the relation between a word's frequency and rank in a corpus, the
distribution of syntactic dependency lengths across languages, or the
presence of properties such as ambiguity. We discuss a framework for
studying how these system-level properties emerge from local,
in-the-moment interactions of rational, pragmatic speakers and
listeners. To do so, we derive a novel objective function for measuring
the communicative efficiency of linguistic systems in terms of the
interactions of speakers and listeners. We examine the behavior of this
objective in a series of simulations focusing on the communicative
function of ambiguity in language. These simulations suggest that
rational pragmatic agents will produce communicatively efficient systems
and that interactions between such agents provide a framework for
examining efficient properties of language structure and use more
broadly.

@@

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007). These projects typically employ
information-theoretic principles to characterize efficiency in natural
or simulated language corpora. We adopt a different approach in this
work, deriving a measure of efficiency from first principles of
\textit{efficient language use} and then showing how this can lead to
communicatively efficient systems.\par

Functionalist theories commonly frame language efficiency in terms of a
fundamental effort-assymetry underlying everyday communication: what is
``hard'' for a speaker is likely different than what is ``hard'' for a
listener. Zipf gave the following intution for how this idea -- purely
from the standpoint of speaker effort, an optimal language
\(\ell_{speaker}^*\) would tend toward a vocabulary of a single,
low-cost word. Given such a language, the full set of potential meanings
would be conveyed using only that word, i.e. \(\ell_{speaker}^*\) would
be fully ambiguous and all possible meanings would need to be
disambiguated by a listener. From the standpoint of listener effort, an
optimal language \(\ell_{listener}^*\) would map all possible meanings
to distinct words, removing a listener's need to disambiguate. In this
example, speaker effort is related to \emph{production cost} and
listener effort to \emph{understanding or disambiguation cost}. Clearly,
natural languages fall between the two extremes of \(\ell_{speaker}^*\)
and \(\ell_{listener}^*\). Zipf proposed that the particular
lexicon-level properties he observed were a result of these competing
forces -- the pressure to jointly minimize speaker and listener
effort.\par

But how does this optimization take place? The example given by Zipf
(1949), describes local, communicative interactions in terms of a
\textit{reference game}. Speakers intend to refer to some object in the
world \(m\). They choose some utterance \(u\) to transmit this intended
meaning, \(u \rightarrow m\). The listener attempts to reconstruct this
intended meaning given the transmitted utterance, \(m \rightarrow u\).
Other projects have assumed this basic reference game setting
(Piantadosi et al., 2011; Regier et al., 2015) and this simplification
of the communicative act has proven productive in theoretical
(Ferrer-i-Cancho, 2018), simulation-based (Kirby et al., 2014) and
empirical explorations (Hawkins, Franke, Smith, \& Goodman, 2018) of
efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to
speakers' pragmatic reasoning in the moment. Under a Gricean treatment
of pragmatics, speakers and listeners follow a set of conversational
maxims in which they cooperate to transfer information (Grice, 1975).
These maxims appear to emerge from efficiency concerns, however (Horn,
1984). We formalize this connection -- showing how system-level
efficiencies can emerge from local interaction behavior. Intuitively,
our claim is that to understand an ``efficient'' property of a system it
is essential that we consider how that property is \emph{used}
efficiently. It follows that any quantitative measure of ``efficiency''
must be a function of both the system and its users.\par

In the current work we provide a case-study for the framing presented
above. We choose a property of languages that could, in principle, vary
freely, but shows strong regularities across languages. The explanadum
is why this regularity holds. We examine ambiguity as our property,
extending ideas by Piantadosi et al. (2011). We define a novel measure
of efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users using the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these basic
ingredients, we show that the property of interest (ambiguity) is
prevalent in languages that optimize our measure of efficiency. Further,
we show how ambiguity is \textit{used} efficiently during local,
in-the-moment interactions. The contributions of this work are twofold
-- we derive a novel objective for measuring linguistic efficiency and
also show how the reference game framework, in combination with formal
models of communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

Zipf's example of optimal speaker- and listener-languages took the form
of a reference game. We adopt that formulation here, assuming these
communication games as our basic unit of analysis. In this framework,
speakers and listeners are aware of a set of objects \(M\)
(\emph{meanings}) and are knowledgeable about the set of possible
signals \(U\) (\emph{utterances}) that can be used to refer to a given
meaning (see Figure 1). Utterances may have different relative costs,
operationalized via a prior over utterances \(P(U)\). Similarly,
meanings differ in the relative degree to which they need to be talked
about, operationalized as a prior over meanings
\(P(M)\)\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work (Regier, Kemp \& Kay (2015).}.
We consider a set of contexts \(c\in C\) with an associated prior
\(P(C)\). Each context \(c_i\) describes a different distribution over
meanings e.g. \(p(M|C=c_i) \neq p(M|C=c_j)\). Finally, we consider a set
of communicative events \(e \in E\) where \(<u, m, c> = e\) is an
utterance, meaning and context triple.\par

@@

A language \(\ell\) defines the set of semantic mappings between
utterance and meanings. For example, Figure 1 contains four utterances
\(U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}\).
The boolean matrix describes the literal semantics of the language. We
define a language as ``ambiguous'' if there is some utterance
\(u \in U\) which can apply to multiple meanings (i.e.
\(|[[u_i]]| > 1\))\footnote{We use double brackets $[[\dots]]$ to represent denotation.}.
In Figure 1 both the words ``square'' and ``green'' are ambigous so we
would say that \(\ell\) contains ambiguity.

@@

The Rational Speech Act framework (RSA) is computational-level theory of
pragmatic language use, which has been well attested in experimental
work examining both speaker and listener communication behavior (Frank
\& Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game
setting\footnote{See SI section 1 for a more detailed description of basic RSA speaker-listener defintions and to Goodman \& Frank [-@GoodmanFrank2016a] for an overview of its applications.}.\par

An RSA \emph{speaker agent} defines a conditional distribution over
utterances, mapping from intended meanings \(M\) to utterances \(U\)
using \(\ell\) in a given context \(c\). That is, a speaker defines
\(P_{speaker}(u|m, c; \ell)\). We will use \(S(u|m, c; \ell)\) as
short-hand throughout. A \emph{listener agent} defines a conditional
distribution over meanings, mapping from utterances \(U\) to meanings
\(M\) using \(\ell\) in a given context \(c\) (i.e. \(L(m|u, c;\ell)\)).
Note that both speakers and listeners can induce joint distributions
over utterace-meaning pairs, although, importantly, these distributions
may differ: \[P_{speaker}(u, m | c; \ell) = S(u|m, c; \ell)p(m|c)\]
\[P_{listener}(u, m| c; \ell) = L(m|u, c; \ell)p(u|c)\]

@@

Zipf (1949) proposed that the particular distributional properties found
in natural language emerge from competing speaker and listener
pressures. We operationalize this objective in equation (1) -- the
efficiency of a linguistic system \(\ell\) being used by speaker and
listener agents \(S\) and \(L\) is the sum of the expected speaker and
listener effort to communicate over all possible communicative events
\(e \in E\).\par

@@

We assume that speaker effort is related to the suprisal of an utterance
in a particular
context\footnote{In the current set of simulations we consider utterances costs as independent from context (ie. $p(u|c)p(c)=p(u)p(c)$).}
-- intuitively, the number of bits needed to encode the utterance \(u\).
This particular formalization of speaker-cost is general enough to
accommodate a range of cost instantiations, such as production difficult
via articulation effort, cognitive effort related to lexical access, or
others (Bennett \& Goodman, 2018).\par

\[\text{speaker effort} = -log_2(p(u|c))\]

We assume listener effort is the semantic suprisal of a meaning given an
utterance. This operationalization of listener effort is intuitively
related to existing work in sentence processing in which word
comprehension difficulty is proportional to surprisal (Hale, 2001; Levy,
2008).

\[\text{listener effort} = -log_2(L(m|u, c; \ell))\]

Importantly, we assume that events \(e = <u, m, c>\) are sampled
according the to following generative model -- some context occurs in
the world with probability \(P(C=c)\). Within this context, an object
\(m\) occurs with probability \(p(m|c)\). The speaker attempts to refer
to that object by sampling from her conditional distribution
\(S(u|m, c; \ell)\) (i.e. \(e \sim p(c)p(m|c)S(u|m, c; \ell)\)). From
the ingredients described above, it is possible to derive the following
objective between the speaker and listener joint distributions (see SI
2.2 for a complete derivation).

\begin{equation}
\begin{split}
  = \mathbb{E}_{c \sim P(C)}[H_{cross}(P_{speaker}, P_{listener} | c; \ell)]\\
\end{split}
\end{equation}

From an information-theoretic perspective this objective is intuitive:
The Cross-Entropy (CE) gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, we have an expectation over this term -- the expected
difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\). In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to
handle various forms of ambiguity: lexical, syntactic, among
others(Wasow, Perfors, \& Beaver, 2005). The ubiquity of this property,
however, has been argued to provide evidence that langauges are not been
optimized for
communication\footnote{Chomsky's (2002) claim that "... language is not well designed, because you have such properties as ambiguity" is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$.}
(Chomsky, 2002).

@@

We re-iterate their argument here. Say we have two objects (\(m_1\) and
\(m_2\)), two utterances (\(u_1\) and \(u_2\)), which differ in cost,
and two languages (\(\ell_1\) and \(\ell_2\)) which describe different
utterance-meaning mappings. In language \(\ell_1\), the low-cost \(u_1\)
can be used to refer to both \(m_1\) and \(m_2\)
(\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\)), but the high-cost \(u_2\) cannot
be used at all (\([[u_2]]_{\ell_1} = \emptyset\)). By contrast, in
language \(\ell_2\), \(u_1\) can only refer to \(m_1\) and \(u_2\) can
only refer to \(m_2\) (\([[u_1]]_{\ell_2} = \{m_1\}\) and
\([[u_2]]_{\ell_2} = \{m_2 \}\)). While it is cheaper for a speaker to
use \(\ell_1\) (because speaking is always low-cost), it is more
difficult for a listener (because \(u_1\) is ambiguous). Crucially,
however, if context is disambiguating then the speaker can use \(u_1\)
to refer to either \(m_1\) or \(m_2\) and \(\ell_1\) should be preferred
to \(\ell_2\).\par

@@

\caption[Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases]{Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages containing ambiguity. Horizontal axis shows the number of context-sizes (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only objectives. Error bars represent 95 percent confidence intervals.}\label{fig:plot-optimal-langs}
\end{figure*}
\end{CodeChunk}

@@

We adopt the hypothesis presented by Piantadosi, formalizing their
argument in our current framework -- showing that ambiguity is an
efficient property under our CE objective in the reference game setting.
We proceed by generating languages with different amounts of contextual
support (varying the size of \(|C|\)). We search the space of languages,
examining whether ones which minimize our objective contain ambiguity.
If context leads to more efficient commnication, then optimal languages
under our objective should be more likley to be ambiguous as the amount
of context increases.\par

@@

We conduct \(N=2000\) simulations. For each simulation we enumerate the
set of \emph{valid} languages in which \(|U|=|M|=4\) (recall that \(U\)
is our set of utterances and \(M\) our set of meanings). Note that a
language \(\ell \in L\) is ``valid'' so long as each possible meaning in
\(m \in M\) can be referred to by at least one form \(u \in U\) (every
column of \(\ell\) has some non-zero assignment) and each form maps to
at least one meaning (every column has some non-zero assignment). For a
given simulation the goal is to find the language \(\ell^*\) which
minimizes our CE objective and then check to see if that language
contains ambiguity.\par

We define language efficiency as a function of the particular semantic
mappings induced by that language, the speaker and listener agents
(\(S\) and \(L\)), as well as the utterance (\(P(U)\)), meaning
(\(P(M)\)), and context priors (\(P(C)\)). Rather than assume particular
structure, for each simulation we generate
\(P(U) \sim \text{Dir}(1, |U|)\), \(P(M|C=c) \sim \text{Dir}(1, |M|)\)
(a separate conditional distribution over meanings for each context
\(c\)), and \(P(C) \sim \text{Dir}(1, |C|)\), where \(\text{Dir}(1, k)\)
specifies the uniform Dirichlet distribution over a \(k\)-dimensional
probability vector.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on the presence of ambiguity in
optimal languages. To do so we consider four conditions with \(n=500\)
simulations each (that is, 500 unique sets of
\(\{P(U), P(M|C), P(C)\}\). Our first is a \emph{one-context} condition
(\(|C|=1\)) -- only a single distribution over meanings \(P(M)\). In our
two-context condition (\(|C| = 2\)), we consider efficiency under both
\(P(M|C=c_1)\) as well as \(P(M|C=c_2)\). Three- and four-context
conditions corresponding accordingly.\par

@@

For comparison, we also examine properties of optimal languages under
two additional objectives. Zipf (1949) proposed that the speaker-optimal
language \(\ell_{speaker}^*\) would minimize speaker effort and the
listener-optimal language \(\ell_{listener}^*\) would minimize listener
effort. We define these objectives using the first and second half of
equation 1 (see SI section 2.2. for more detail).\par

@@

Figure 1. plots the proportion of optimal languages under each objective
as a function of context size. The red line shows that as the number of
contexts increases, so does the probability that an optimal language
\(\ell^*_{cross}\) contains ambiguity under our CE objective. For
comparison we also plot the proportion of speaker-optimal
\(\ell^*_{speaker}\) (blue line) and listener-optimal
\(\ell^*_{listener}\) (green line) languages that contain ambiguity. In
line with Zipf's predictions, if languages are designed only to minimize
speaker effort then optimal languages always contain ambiguity. If
languages are designed to minimize listener effort then ambiguity is
always avoided.\par

In simulation 1 we explored the degree to which ambiguity was an
efficient property when communication was contextualized. We showed that
as contextual information increased, the optimal languages under our CE
objective were more likely to have ambiguous lexical items. Importantly,
we did not see this effect under alternative objectives which only
emphasized speaker- or listener effort alone.\par

While our results indicate that ambiguity is an efficient property of
contextualized language, these simulations assumed that agents could
always disambiguate items in context. That is, in our four conditions
both agents had perfect knowledge of the relevant conditional
distributions (\(P(M|C)\)). This assumption may be too strong for
describing much of day-to-day communication -- we seldom interact with
others with perfect knowledge of the current context (or topic) of
conversation. To explore how ambiguity may be \textit{used} efficiencly,
we next explore a case in which the listener has imperfect knowledge of
context at the start of the conversation, but may infer it from a
discourse history in Simulation 2.\par

@@

We shift focus in Simulation 2 -- instead of examining the landscape of
a space of languages under our CE objective, we consider a single
language containing ambiguity, examining how that property is used over
a discourse. If context is not informative with respect to meaning early
in a discourse, but is at later positions, then an efficient speaking
strategy would avoid ambiguous material early, saving it for when
contextual information is disambiguating. In this toy example we can
consider ``context'' as analogous to a ``topic'' of conversation. As an
example, imagine a scenario in which a reader is beginning a magazine
article. While they may have some knowledge about the article's topic
(perhaps from the title), they may not have complete knowledge of its
contents, including the persons or events involved. In this setting,
using an potential low-cost, but ambiguous pronoun early may lead to
misunderstanding if context is sufficiently informative. But, if by a
later position enough contextual information has accumulated, it may be
efficient for the writer to use the ambiguous material.\par

@@

\hypertarget{ref-FrankGoodman2012a}{}
Frank, M., \& Goodman, N. (2012). Predicting pragmatic reasoning in
language games. \emph{Science}, \emph{336}, 998.
